Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-31 Manuscript/manuscript.Rmd
edits

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].

A growing number of studies have proposed computational tools used to identify the complete set of key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions [@mourad2016computational; @huang2015predicting; @hong2017computational]. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Additionally, identifying the most informative feature engineering steps used to quantify the relationships between genomic annotations and bins of varying resolutions remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model, whereby loops are formed during interphase [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological]. The growing number of publically available genomic annotation data and their distinct patterns prompt the use of computational approaches toward TAD boundary prediction.

The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions <!--[@mourad2016computational; @huang2015predicting; @hong2017computational]-->. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the hundreds of kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation and its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Additionally, identifying the most informative feature engineering steps used to quantify the relationships between genomic annotations and bins of varying resolutions remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, class imbalance was not addressed and instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any transcription factors, resulting in a zero inflated model. This, likely being further exacerbated by the class imbalance. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS). However, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. There was also a consensus on feature engineering where both methods used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions.
